---
title: "T-testing Example"
author: "D2K Course Staff"
date: "`r format(Sys.time(), '%B %d %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr) # We need the knitr package to set chunk options
library(tidyverse)

# Set default knitr options for knitting code into the report:
opts_chunk$set(echo=TRUE,  # change to FALSE to keep code out of the knitted document
               cache=FALSE, # re-run code that has already been run?
               autodep=TRUE, # assure that caching dependencies are updated correctly
               cache.comments=FALSE, # do not re-run a chunk if only comments are changed
               message=FALSE, # change to FALSE to keep messages out of the knitted document
               warning=FALSE,  # change to FALSE to keep warnings out of the knitted document
               comment = NA,
               tidy.opts=list(width.cutoff=65))

theme1 <- theme_bw() +
  theme(axis.text = element_text(size = 8, colour = "#6b3447"),
        axis.title = element_text(size = 10, colour = "#2f2f63"),
        legend.title = element_text(size = 8, colour = "#2f2f63"),
        legend.text = element_text(size = 8, colour = "#6b3447"), 
        title = element_text(size = 12, colour = "#2f2f63"), 
        axis.ticks = element_line(colour = "#6b3447"),
        plot.caption = element_text(size = 8, colour = "#2f2f63"),
        plot.subtitle = element_text(size = 10, colour = "#2f2f63"))
```

# Data

Below, we show an example of how to run two sample t-tests as well as two sample proportion tests. This is done on the `iris` data set, which contains data on 3 different species of iris flowers. 

```{r}
data(iris)
head(iris)
```

In this data set, `Species` is a categorical variable, while the sepal and petal lengths and widths are continuous variables. For this data set, we will be comparing the `setosa` and `viginica` species, so we create subsets of the data:

```{r}
library(dplyr)

iris_setosa <- dplyr::filter(iris, Species == 'setosa')
iris_virginica <- dplyr::filter(iris, Species == 'virginica')
```


\newpage

# Two-Sample T-test

The two-sample t-test is used to compare the means of two groups of samples. For example, say we are interested in testing if the mean sepal length is the same for the `setosa` and `viginica` species in the `iris` data set. Formally, we are running a hypothesis test with:

$H_0$: The mean sepal lengths for the `setosa` and `viginica` species are the same.

$H_a$: The mean sepal lengths for the `setosa` and `viginica` species are different.

The assumptions for the t-test are:

\begin{itemize}
    \item The population follows a normal distribution.
    \item Samples are independent.
    \item The samples are representative of the population.
\end{itemize} 

The two-sample t-test can be run under the assumption that the population variances are equal between the two groups, or not. (The former is also known as a "pooled variance" test while the latter is known as an "unpooled variance" test.) The results of both are shown below: 

```{r}
# Equal variance
t.test(iris_setosa$Sepal.Length, iris_virginica$Sepal.Length,
       var.equal = TRUE)

# Unequal variance
t.test(iris_setosa$Sepal.Length, iris_virginica$Sepal.Length,
       var.equal = FALSE)
```

Notice that the confidence intervals are slightly different between the two tests. However, in both of these cases, the listed p-value is extremely small, meaning that there is a significant difference between the sepal lengths for the `setosa` and `viginica` species. (Note: in the real world, only one test should be chosen and should be done so before the test is run.) 

\newpage

# Two-Sample Proportion Test

The two-sample proportion test is used to compare the proportions of a binary categorical variable of two groups of samples. For example, say we want to test the hypothesis:

$H_0$: The proportions of sepal widths greater than 3 cm for the `setosa` and `viginica` species are the same.

$H_a$: The proportions of sepal widths greater than 3 cm for the `setosa` and `viginica` species are different.

The assumptions for two-sample proportions test are:

\begin{itemize}
    \item Samples are not a substantial proportion of the entire population for both groups.
    \item Samples are independent.
    \item There are a sufficient number of successes and failures in each group.
\end{itemize} 

First, we have to calculate the propotions:

```{r}
# Calculate proportions
prop.test(c(sum(iris_setosa$Sepal.Width > 3), sum(iris_virginica$Sepal.Width > 3)), 
          c(nrow(iris_setosa), nrow(iris_virginica)))
```

The listed p-value is extremely small, meaning that there is a significant difference between the proportions of flower sepal widths that are above 3 cm for the `setosa` and `viginica` species.